---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fancyfx

<!-- badges: start -->
[![R-CMD-check](https://github.com/chross22/fancyfx/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/chross22/fancyfx/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

An effect curve on its own is easy to over-read. A model will happily draw a
confident-looking bend at the far right of the x axis, and nothing in the plot
tells you that only three observations sit under it.

`fancyfx` pairs every effect curve with a rug of the raw data, drawn directly
above it on a shared x axis, so the shape of the effect and the weight of
evidence behind it get read together.

The second thing it is for is **getting that figure into a manuscript without a
further round of fiddling**. Defaults are chosen for publication rather than for
exploration — a clean theme with no grid or background panel, lettered panel
labels, and a categorical palette checked for legibility under colour vision
deficiency — so a bare call gets you close to the figure you would submit. Every
one of those is an argument, so a house style can replace any of them.

It works across model types: GAMs fitted with `mgcv` are shown as partial
effects via `gratia`, and everything else — including mixed and Bayesian fits —
as predictions via `marginaleffects`.

Alongside the effect plots are **model evaluation plots** — ROC/AUC, the
TSS-versus-threshold trade-off, and permutation importance — which ask whether
the model earns the effects it reports.

> **Upgrading?** The package used to handle only GAMs, and was named for it.
> `plotSmooths()` still works and produces an identical plot, but is deprecated
> in favour of `plotEffects()` — same arguments, same output. See
> [Migrating](#migrating-from-plotsmooths).

## Installation

You can install the development version of fancyfx from
[GitHub](https://github.com/chross22/fancyfx) with:

``` r
# install.packages("devtools")
devtools::install_github("chross22/fancyfx")
```

## Example

```{r fancyfx-gam}
library(fancyfx)

# A GAM, using the iris data set available in R
gam.fit <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)

plotEffects(gam.fit, iris, "Sepal.Length", xlab = "Sepal length (cm)")
```

The histogram along the top is the point: where it is thin, be careful.

Several terms can be shown at once, each keeping its own rug:

```{r fancyfx-multi}
gam.fit2 <- mgcv::gam(Petal.Length ~ s(Sepal.Length) + s(Petal.Width),
                      data = iris)

combinePlots(gam.fit2, iris, vars = c("Sepal.Length", "Petal.Width"),
             title = "Partial effects on petal length")
```

Non-GAM models take exactly the same call. Here a logistic regression, on the
scale of the outcome, with a 95% interval:

```{r fancyfx-glm}
glm.fit <- glm(am ~ wt + hp, data = mtcars, family = binomial)

plotEffects(glm.fit, mtcars, "wt",
            scale = "response", interval = "ci",
            xlab = "Weight (1000 lbs)",
            ylab = "P(manual transmission)")
```

## Comparing models

`comparePlots()` holds the variable fixed and varies the *model*, which is how
you check whether a modelling choice bought you anything. A factor-smooth
interaction is drawn as one curve per level, with a colourblind-safe palette
and a legend:

```{r fancyfx-compare, fig.height = 4}
plain <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)
by.species <- mgcv::gam(Petal.Length ~ s(Sepal.Length, by = Species) + Species,
                        data = iris)

comparePlots(list("Single smooth" = plain,
                  "Smooth by species" = by.species),
             iris, "Sepal.Length",
             title = "Is a factor-smooth interaction worth it?")
```

## Rug styles

`rug.type` picks how the raw data is summarised above the curve. A histogram
shows counts and reads well at moderate sample sizes; a density is smoother and
works better when a histogram would be noisy.

```{r fancyfx-rugs, fig.height = 4}
lm.fit <- lm(mpg ~ wt + hp, data = mtcars)

ggpubr::ggarrange(
  plotEffects(lm.fit, mtcars, "wt", rug.type = "histogram",
              xlab = "Weight (1000 lbs)"),
  plotEffects(lm.fit, mtcars, "wt", rug.type = "density",
              xlab = "Weight (1000 lbs)"),
  labels = c("A", "B")
)
```

## Publication-ready by default

Plots use `theme_fancyfx()`, built on `ggpubr::theme_pubr()`: no background
panel, no grid, plain axis lines, and text sized to survive being shrunk into a
column. Panels are labelled `A`, `B`, `C` by default.

```r
# Bigger text for a narrow figure
plotEffects(fit, dat, "x", theme = theme_fancyfx(base_size = 14))

# Lower-case, numbered, none, or your own
combinePlots(fit, dat, vars, labels = "a")
combinePlots(fit, dat, vars, labels = "1")
combinePlots(fit, dat, vars, labels = "none")
combinePlots(fit, dat, vars, labels = c("Panel one", "Panel two"))

# Any other ggplot2 theme works too
plotEffects(fit, dat, "x", theme = ggplot2::theme_minimal())
```

Curves that split by a factor use `fancyfx_palette()`, a six-colour categorical
palette chosen by search rather than by eye: every colour sits in a mid
lightness band, clears 3:1 contrast against a white page, and stays separable
under simulated protanopia and deuteranopia.

## Evaluating a model

Effect plots say what a model claims. These say whether to believe it.

`plotROC()` and `plotThreshold()` cover presence/absence models — the
discrimination and cutoff questions respectively — and `plotImportance()` covers
which predictors the model is actually leaning on.

```{r fancyfx-eval, fig.height = 4}
set.seed(1)
d <- data.frame(x1 = runif(600, 1, 10), x2 = runif(600, 1, 10),
                x3 = runif(600, 1, 10))
d$y <- rbinom(600, 1, plogis(-3 + 0.6 * d$x1))
train <- d[1:300, ]
test  <- d[301:600, ]

sdm <- glm(y ~ x1 + x2 + x3, data = train, family = binomial)

ggpubr::ggarrange(
  plotROC(sdm, test),
  plotImportance(sdm, test, n.perm = 20),
  labels = c("A", "B"), widths = c(1, 1.2)
)
```

A ROC curve says how well the model *ranks*; it does not tell you where to cut.
`plotThreshold()` does — sensitivity and specificity against the cutoff, with
the TSS-maximising threshold marked:

```{r fancyfx-threshold, fig.height = 4}
plotThreshold(sdm, test)
```

### Evaluation data is required, on purpose

`newdata` is a required argument, and passing the training data warns and
annotates the figure as in-sample. This is deliberate: an in-sample ROC can look
excellent for a model with no predictive value at all, and the easiest figure to
produce should not be the misleading one.

Cross-validation folds are supported via `folds`, drawn per fold so the spread
is visible rather than averaged away — but they come with a note, because
cross-validated metrics are weaker evidence than an independent hold-out. For
spatial models the gap is wider still: with random folds a held-out point
usually has a near neighbour among the training folds, so the model has in
effect already seen it. Use spatially blocked folds.

```r
plotROC(sdm, test, folds = spatial_block_ids)
```

Two caveats the functions state in their own documentation, worth repeating:
AUC and TSS are defined for **binary outcomes only** and a continuous response
is refused rather than scored; and permutation importance **splits credit badly
between correlated predictors**, which bites hard on environmental covariates.

## Supported models

| Model | Backend | What you get |
|---|---|---|
| `mgcv::gam()`, `bam()` | `gratia` | Partial effect, link scale, centered |
| `lm()`, `glm()` | `marginaleffects` | Predicted values |
| `lme4::lmer()`, `glmer()` | `marginaleffects` | Predicted values, population level |
| `glmmTMB::glmmTMB()` | `marginaleffects` | Predicted values, population level |
| `brms::brm()` | `marginaleffects` | Predicted values, credible interval |
| `rstanarm::stan_glm()`, `stan_glmer()` | `marginaleffects` | Predicted values, credible interval |
| Most other fitted models | `marginaleffects` | Predicted values |

Support beyond those comes from whatever `marginaleffects` handles. The rows
above are the families verified against real fits.

**Bayesian fits** are summarised from posterior draws, which give an interval
but no standard error — so the ribbon is the credible interval at `level`, and
there is no `±1 SE` ribbon to be had. Write `interval = "cri"` if you would
rather say so explicitly; it computes the same thing. `brms` takes `re_formula`
rather than `re.form`, and that translation is handled for you.

For a **mixed model**, `re.form` defaults to `NA`, so the effect is drawn at the
population level. This matters: left to the backend's own default, the grouping
factor is held at its modal level and the plot silently shows the effect for one
arbitrary group. Note that the ribbon covers uncertainty in the fixed effects
only.

### A caveat worth reading

The two backends compute **different quantities**. A partial effect is one
term's contribution in isolation, centered so it averages to zero. A predicted
value is the model's fitted output with the other predictors held at
representative values. `fancyfx` labels the y axis with whichever it computed,
and you should not compare them as though they were on the same footing.

The default `interval = "se"` ribbon spans roughly 68%, not 95% — it is this
package's historical default, kept so existing figures do not silently change.
Pass `interval = "ci"` for a conventional confidence interval.

See `vignette("fancyfx")` for the full discussion.

## Migrating from plotSmooths()

```r
# Old:
plotSmooths(gam.fit, iris, "Sepal.Length")

# New — same arguments, same output:
plotEffects(gam.fit, iris, "Sepal.Length")
```

The defaults reproduce exactly what `plotSmooths()` drew for a GAM, so
migrating is a rename and nothing more.

## How to cite

```r
citation("fancyfx")
```

`fancyfx` is a plotting layer over other people's estimation work. Please also
cite whichever package computed your effects: `gratia` and `mgcv` for GAM
partial effects, `marginaleffects` otherwise.
