---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fancyfx

<!-- badges: start -->
[![R-CMD-check](https://github.com/chross22/fancyfx/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/chross22/fancyfx/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

An effect curve on its own is easy to over-interpret. A model could draw a
confident-looking bend at the far right of the x axis, and nothing in the plot
would tell you that only three observations sit under it.

`fancyfx` pairs every effect curve with a rug of the raw data, drawn directly
above it on a shared x axis, so the shape of the effect and the weight of
evidence behind it get read together. Additionally, `fancyfx` returns manuscript
ready figures without having to fiddle with settings. Defaults are chosen for
publication rather than for exploration — a clean theme with no grid or background panel, lettered panel
labels, and a categorical palette checked for legibility under colour vision
deficiency — so a bare call gets you close to the figure you would submit. Every
one of those is an argument, so a house style can replace any of them.

It works across model types: GAMs fitted with `mgcv` are shown as partial
effects via `gratia`, and everything else — including mixed and Bayesian fits —
as predictions via `marginaleffects`.

Alongside the effect plots are **model evaluation plots** — ROC/AUC, the
TSS-versus-threshold trade-off, and permutation importance — which ask whether
the model earns the effects it reports.

## Which function do I want?

| I want to… | Use |
|---|---|
| plot one predictor's effect | `plotEffects()` |
| plot several predictors | `combinePlots()` |
| compare competing models | `comparePlots()` |
| get the effect numbers, not a plot | `effect_estimates()` |
| draw a rug on its own | `plotRugs()` |
| **evaluate a presence/absence model** | |
| how well does it rank? | `plotROC()` |
| where should the cutoff go? | `plotThreshold()` |
| are the probabilities honest? | `plotCalibration()` |
| which predictors is it using? | `plotImportance()` |
| how much deviance is explained? | `calc_deviance()` |
| is my hold-out actually independent? | `spatial_sorting_bias()` |
| score predictions I already have (CV) | `held_out()` |
| **spatial projections** | |
| where does the ensemble disagree? | `plotUncertainty()` |
| where is it extrapolating? | `plotExtrapolation()` |
| aggregate to a defensible resolution | `plotHexbin()` / `hex_bin()` |
| thin records clustered by survey effort | `thin_points()` |
| do two distributions overlap? | `niche_overlap()` / `niche_equivalency()` |
| **make it publication-ready** | |
| change the theme or font sizes | `theme_fancyfx()` |
| change the curve colours | `fancyfx_palette()` |

## Installation

You can install the development version of fancyfx from
[GitHub](https://github.com/chross22/fancyfx) with:

``` r
# install.packages("devtools")
devtools::install_github("chross22/fancyfx")
```

## Example

```{r fancyfx-gam}
library(fancyfx)

# A GAM, using the iris data set available in R
gam.fit <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)

plotEffects(gam.fit, iris, "Sepal.Length", xlab = "Sepal length (cm)")
```

The histogram along the top is the point: where it is thin, be careful.

Several terms can be shown at once, each keeping its own rug:

```{r fancyfx-multi}
gam.fit2 <- mgcv::gam(Petal.Length ~ s(Sepal.Length) + s(Petal.Width),
                      data = iris)

combinePlots(gam.fit2, iris, vars = c("Sepal.Length", "Petal.Width"),
             title = "Partial effects on petal length")
```

Non-GAM models take exactly the same call. Here a logistic regression, on the
scale of the outcome, with a 95% interval:

```{r fancyfx-glm}
glm.fit <- glm(am ~ wt + hp, data = mtcars, family = binomial)

plotEffects(glm.fit, mtcars, "wt",
            scale = "response", interval = "ci",
            xlab = "Weight (1000 lbs)",
            ylab = "P(manual transmission)")
```

## Comparing models

`comparePlots()` holds the variable fixed and varies the *model*, which is how
you check whether a modelling choice bought you anything. A factor-smooth
interaction is drawn as one curve per level, with a colourblind-safe palette
and a legend:

```{r fancyfx-compare, fig.height = 4}
plain <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)
by.species <- mgcv::gam(Petal.Length ~ s(Sepal.Length, by = Species) + Species,
                        data = iris)

comparePlots(list("Single smooth" = plain,
                  "Smooth by species" = by.species),
             iris, "Sepal.Length",
             title = "Is a factor-smooth interaction worth it?")
```

## Rug styles

`rug.type` picks how the raw data is summarised above the curve. A histogram
shows counts and reads well at moderate sample sizes; a density is smoother and
works better when a histogram would be noisy.

```{r fancyfx-rugs, fig.height = 4}
lm.fit <- lm(mpg ~ wt + hp, data = mtcars)

ggpubr::ggarrange(
  plotEffects(lm.fit, mtcars, "wt", rug.type = "histogram",
              xlab = "Weight (1000 lbs)"),
  plotEffects(lm.fit, mtcars, "wt", rug.type = "density",
              xlab = "Weight (1000 lbs)"),
  labels = c("A", "B")
)
```

## Publication-ready by default

Plots use `theme_fancyfx()`, built on `ggpubr::theme_pubr()`: no background
panel, no grid, plain axis lines, and text sized to survive being shrunk into a
column. Panels are labelled `A`, `B`, `C` by default.

```r
# Bigger text for a narrow figure — scales every element together
plotEffects(fit, dat, "x", theme = theme_fancyfx(base_size = 16))

# Or size each element on its own
plotEffects(fit, dat, "x",
            theme = theme_fancyfx(base_size = 13,
                                  axis.title.size = 18,
                                  axis.text.size = 10,
                                  legend.title.size = 15))

# Panel labels and the figure title are drawn by the arranging step,
# so they have their own arguments
combinePlots(fit, dat, vars, title = "...", label.size = 20, title.size = 18)

# Lower-case, numbered, none, or your own
combinePlots(fit, dat, vars, labels = "a")
combinePlots(fit, dat, vars, labels = "1")
combinePlots(fit, dat, vars, labels = "none")
combinePlots(fit, dat, vars, labels = c("Panel one", "Panel two"))

# Any other ggplot2 theme works too
plotEffects(fit, dat, "x", theme = ggplot2::theme_minimal())
```

Curves that split by a factor use `fancyfx_palette()`, a six-colour categorical
palette chosen by search rather than by eye: every colour sits in a mid
lightness band, clears 3:1 contrast against a white page, and stays separable
under simulated protanopia and deuteranopia.

## Evaluating a model

Effect plots say what a model claims. These say whether to believe it.

For presence/absence models, `plotROC()` covers discrimination,
`plotThreshold()` covers where to cut, and `plotCalibration()` covers whether
the probabilities are honest. `plotImportance()` covers which predictors the
model is actually leaning on, for any model type.

```{r fancyfx-eval, fig.height = 4}
set.seed(1)
d <- data.frame(x1 = runif(600, 1, 10), x2 = runif(600, 1, 10),
                x3 = runif(600, 1, 10))
d$y <- rbinom(600, 1, plogis(-3 + 0.6 * d$x1))
train <- d[1:300, ]
test  <- d[301:600, ]

sdm <- glm(y ~ x1 + x2 + x3, data = train, family = binomial)

ggpubr::ggarrange(
  plotROC(sdm, test),
  plotImportance(sdm, test, n.perm = 20),
  labels = c("A", "B"), widths = c(1, 1.2)
)
```

A ROC curve says how well the model *ranks*; it does not tell you where to cut.
`plotThreshold()` does — sensitivity and specificity against the cutoff, with
the TSS-maximising threshold marked:

```{r fancyfx-threshold, fig.height = 4}
plotThreshold(sdm, test)
```

And neither says whether the probabilities themselves are *honest*.
`plotCalibration()` does: a model that says 0.7 should be right about 70% of
the time.

```{r fancyfx-calibration, fig.height = 5}
plotCalibration(sdm, test)
```

Discrimination and calibration are genuinely separate questions, and AUC cannot
answer the second: it only cares about ranking, so it is unchanged by any
monotone rescaling — a model can post an excellent AUC while every probability
it reports is far too extreme. The reported slope makes it concrete: 1 is
perfect, below 1 means over-confident.

Note the rug. Calibration is usually worst at the extremes, and the extremes
usually hold the fewest predictions, so the most eye-catching departures from
the diagonal are often the least trustworthy points on the plot.

Three defaults here are deliberate. **`newdata` is required**, and passing the
training data warns and annotates the figure as in-sample — an in-sample ROC can
look excellent for a model with no predictive value, and the easiest figure to
produce should not be the misleading one. **Folds are drawn per fold** rather
than averaged, and come with a note, because cross-validated metrics are weaker
evidence than an independent hold-out — for spatial models, use spatially
blocked folds. And **`spatial_sorting_bias()`** says how far a split falls short
of independence: near 1 it is doing its job, near 0 the test presences sit so
close to the training data that AUC is measuring the split rather than the
species.

Two limits the functions state themselves: AUC and TSS are defined for **binary
outcomes only**, and permutation importance **splits credit badly between
correlated predictors**, which bites hard on environmental covariates.

→ [**Evaluating a model**](vignettes/evaluation.Rmd) covers all of it, plus
deviance, variable importance, and how to put it in a paper.

## Spatial projections

A projection map is a persuasive object. It fills the study area with colour,
looks identical whether the model had a thousand observations in a region or
none, and nothing on it separates the part built on evidence from the part built
on the model's willingness to keep predicting.

Two maps put that distinction back. `terra` is a suggested package, so nothing
here is installed for users who never project.

```{r fancyfx-spatial, eval = requireNamespace("terra", quietly = TRUE), fig.height = 3.6}
library(terra)

set.seed(1)
grid <- rast(nrows = 50, ncols = 65, xmin = -71, xmax = -65,
             ymin = 41, ymax = 45, crs = "EPSG:4326")
lon <- init(grid, "x"); lat <- init(grid, "y")

sst <- 14 - 1.2 * (lat - 41) + 0.3 * (lon + 71); names(sst) <- "sst"
depth <- 20 + 30 * (lon + 71) + 25 * (45 - lat); names(depth) <- "depth"
covariates <- c(sst, depth)

# A survey covering only the north-west of the domain
points <- data.frame(lon = runif(400, -71, -67.5), lat = runif(400, 42, 45))
survey <- cbind(points, extract(covariates, points, ID = FALSE))
survey$present <- rbinom(400, 1,
                         plogis(-4 + 0.45 * survey$sst - 0.01 * survey$depth))

# A bootstrap ensemble of projections
ensemble <- rast(lapply(1:6, function(i) {
  refit <- mgcv::gam(present ~ s(sst) + s(depth),
                     data = survey[sample(nrow(survey), replace = TRUE), ],
                     family = binomial)
  terra::predict(covariates, refit, type = "response")
}))

ggpubr::ggarrange(
  plotUncertainty(ensemble, title = "Ensemble disagreement"),
  plotExtrapolation(covariates, survey, title = "Novel conditions"),
  labels = c("A", "B")
)
```

The south-east is both where the ensemble disagrees most **and** where the
projection has left the surveyed envelope — the honest reading being that the
model has nothing to say about it.

`plotExtrapolation()` draws a MESS surface: below zero, a cell is outside the
training range of at least one covariate. It works one covariate at a time, so
it cannot see novel *combinations* of individually ordinary values — treat a
clean surface as the absence of one specific problem, not permission to project.

Also here: `hex_bin()` and `plotHexbin()` aggregate a raster or point data into a
hexagonal lattice; `thin_points()` thins records where clustering reflects survey
effort rather than the species; and `niche_overlap()` with `niche_equivalency()`
compare two predicted distributions against a randomisation null.

→ [**Spatial projections**](vignettes/spatial.Rmd) covers large rasters,
hexagonal binning, thinning uneven effort, and comparing two distributions.

## Supported models

| Model | Backend | What you get |
|---|---|---|
| `mgcv::gam()`, `bam()` | `gratia` | Partial effect, link scale, centered |
| `gamm4::gamm4()`, `mgcv::gamm()` | `gratia` | Partial effect, population level |
| `scam::scam()` | `gratia` | Partial effect, shape constraint preserved |
| `lm()`, `glm()` | `marginaleffects` | Predicted values |
| `lme4::lmer()`, `glmer()` | `marginaleffects` | Predicted values, population level |
| `glmmTMB::glmmTMB()` | `marginaleffects` | Predicted values, population level |
| `brms::brm()` | `marginaleffects` | Predicted values, credible interval |
| `rstanarm::stan_glm()`, `stan_glmer()` | `marginaleffects` | Predicted values, credible interval |
| Most other fitted models | `marginaleffects` | Predicted values |

Support beyond those comes from whatever `marginaleffects` handles; the rows
above are the families verified against real fits.

The two backends compute **different quantities**, and this is the caveat worth
reading. A partial effect is one term's contribution in isolation, centered so it
averages to zero. A predicted value is the model's fitted output with the other
predictors held at representative values. `fancyfx` labels the y axis with
whichever it computed, and they should not be compared as though they were on the
same footing.

Everything in the GAM family reports a partial effect by default, so those rows
are comparable with each other — `scam` and `gamm4`/`gamm` only because the
package intervenes to unwrap them. Bayesian fits are summarised from posterior
draws, so the ribbon is a credible interval and there is no `±1 SE` to be had.
For a mixed model `re.form` defaults to `NA`, so the effect is drawn at the
population level rather than for one arbitrary group.

The default ribbon is a **95% pointwise interval**, matching `mgcv::plot.gam()`
and `gratia::draw()`. Until 0.10.0 this package drew ±1 SE — roughly 68%, half
the width, and a width most readers would assume was 95%. Pass `interval = "se"`
for that narrower band, or `interval = "simultaneous"` on a GAM smooth.

→ [**Getting started**](vignettes/fancyfx.Rmd) has the full discussion.

## Migrating from plotSmooths()

```r
# Old:
plotSmooths(gam.fit, iris, "Sepal.Length")

# New — same arguments, same output:
plotEffects(gam.fit, iris, "Sepal.Length")
```

The defaults reproduce exactly what `plotSmooths()` drew for a GAM, so
migrating is a rename and nothing more.

## Why it does what it does

Several defaults here are deliberate rather than conventional — evaluation data
being required, `re.form = NA` for mixed models, `na.rm = FALSE` when
summarising an ensemble, the choice of a 95% ribbon over `±1 SE`. Each has a reason,
and the reasons are gathered in [DECISIONS.md](DECISIONS.md) with the
measurements behind them, so a choice can be looked up and defended without
hunting through help pages.

## Documentation

| | |
|---|---|
| [Getting started](vignettes/fancyfx.Rmd) | effect plots, the backends, and what the ribbon means |
| [Evaluating a model](vignettes/evaluation.Rmd) | ROC, thresholds, calibration, deviance, importance |
| [Spatial projections](vignettes/spatial.Rmd) | uncertainty and extrapolation maps, hexbins, thinning |

## How to cite

```r
citation("fancyfx")
```

`fancyfx` is a plotting layer over other people's estimation work. Please also
cite whichever package computed your effects: `gratia` and `mgcv` for GAM
partial effects, `marginaleffects` otherwise.
