---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fancyfx

<!-- badges: start -->
[![R-CMD-check](https://github.com/chross22/fancyfx/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/chross22/fancyfx/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

An effect curve on its own is easy to over-read. A model will happily draw a
confident-looking bend at the far right of the x axis, and nothing in the plot
tells you that only three observations sit under it.

`fancyfx` pairs every effect curve with a rug of the raw data, drawn directly
above it on a shared x axis, so the shape of the effect and the weight of
evidence behind it get read together.

The second thing it is for is **getting that figure into a manuscript without a
further round of fiddling**. Defaults are chosen for publication rather than for
exploration — a clean theme with no grid or background panel, lettered panel
labels, and a categorical palette checked for legibility under colour vision
deficiency — so a bare call gets you close to the figure you would submit. Every
one of those is an argument, so a house style can replace any of them.

It works across model types: GAMs fitted with `mgcv` are shown as partial
effects via `gratia`, and everything else — including mixed and Bayesian fits —
as predictions via `marginaleffects`.

Alongside the effect plots are **model evaluation plots** — ROC/AUC, the
TSS-versus-threshold trade-off, and permutation importance — which ask whether
the model earns the effects it reports.

> **Upgrading?** The package used to handle only GAMs, and was named for it.
> `plotSmooths()` still works and produces an identical plot, but is deprecated
> in favour of `plotEffects()` — same arguments, same output. See
> [Migrating](#migrating-from-plotsmooths).

## Installation

You can install the development version of fancyfx from
[GitHub](https://github.com/chross22/fancyfx) with:

``` r
# install.packages("devtools")
devtools::install_github("chross22/fancyfx")
```

## Example

```{r fancyfx-gam}
library(fancyfx)

# A GAM, using the iris data set available in R
gam.fit <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)

plotEffects(gam.fit, iris, "Sepal.Length", xlab = "Sepal length (cm)")
```

The histogram along the top is the point: where it is thin, be careful.

Several terms can be shown at once, each keeping its own rug:

```{r fancyfx-multi}
gam.fit2 <- mgcv::gam(Petal.Length ~ s(Sepal.Length) + s(Petal.Width),
                      data = iris)

combinePlots(gam.fit2, iris, vars = c("Sepal.Length", "Petal.Width"),
             title = "Partial effects on petal length")
```

Non-GAM models take exactly the same call. Here a logistic regression, on the
scale of the outcome, with a 95% interval:

```{r fancyfx-glm}
glm.fit <- glm(am ~ wt + hp, data = mtcars, family = binomial)

plotEffects(glm.fit, mtcars, "wt",
            scale = "response", interval = "ci",
            xlab = "Weight (1000 lbs)",
            ylab = "P(manual transmission)")
```

## Comparing models

`comparePlots()` holds the variable fixed and varies the *model*, which is how
you check whether a modelling choice bought you anything. A factor-smooth
interaction is drawn as one curve per level, with a colourblind-safe palette
and a legend:

```{r fancyfx-compare, fig.height = 4}
plain <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)
by.species <- mgcv::gam(Petal.Length ~ s(Sepal.Length, by = Species) + Species,
                        data = iris)

comparePlots(list("Single smooth" = plain,
                  "Smooth by species" = by.species),
             iris, "Sepal.Length",
             title = "Is a factor-smooth interaction worth it?")
```

## Rug styles

`rug.type` picks how the raw data is summarised above the curve. A histogram
shows counts and reads well at moderate sample sizes; a density is smoother and
works better when a histogram would be noisy.

```{r fancyfx-rugs, fig.height = 4}
lm.fit <- lm(mpg ~ wt + hp, data = mtcars)

ggpubr::ggarrange(
  plotEffects(lm.fit, mtcars, "wt", rug.type = "histogram",
              xlab = "Weight (1000 lbs)"),
  plotEffects(lm.fit, mtcars, "wt", rug.type = "density",
              xlab = "Weight (1000 lbs)"),
  labels = c("A", "B")
)
```

## Publication-ready by default

Plots use `theme_fancyfx()`, built on `ggpubr::theme_pubr()`: no background
panel, no grid, plain axis lines, and text sized to survive being shrunk into a
column. Panels are labelled `A`, `B`, `C` by default.

```r
# Bigger text for a narrow figure — scales every element together
plotEffects(fit, dat, "x", theme = theme_fancyfx(base_size = 16))

# Or size each element on its own
plotEffects(fit, dat, "x",
            theme = theme_fancyfx(base_size = 13,
                                  axis.title.size = 18,
                                  axis.text.size = 10,
                                  legend.title.size = 15))

# Panel labels and the figure title are drawn by the arranging step,
# so they have their own arguments
combinePlots(fit, dat, vars, title = "...", label.size = 20, title.size = 18)

# Lower-case, numbered, none, or your own
combinePlots(fit, dat, vars, labels = "a")
combinePlots(fit, dat, vars, labels = "1")
combinePlots(fit, dat, vars, labels = "none")
combinePlots(fit, dat, vars, labels = c("Panel one", "Panel two"))

# Any other ggplot2 theme works too
plotEffects(fit, dat, "x", theme = ggplot2::theme_minimal())
```

Curves that split by a factor use `fancyfx_palette()`, a six-colour categorical
palette chosen by search rather than by eye: every colour sits in a mid
lightness band, clears 3:1 contrast against a white page, and stays separable
under simulated protanopia and deuteranopia.

## Evaluating a model

Effect plots say what a model claims. These say whether to believe it.

For presence/absence models, `plotROC()` covers discrimination,
`plotThreshold()` covers where to cut, and `plotCalibration()` covers whether
the probabilities are honest. `plotImportance()` covers which predictors the
model is actually leaning on, for any model type.

```{r fancyfx-eval, fig.height = 4}
set.seed(1)
d <- data.frame(x1 = runif(600, 1, 10), x2 = runif(600, 1, 10),
                x3 = runif(600, 1, 10))
d$y <- rbinom(600, 1, plogis(-3 + 0.6 * d$x1))
train <- d[1:300, ]
test  <- d[301:600, ]

sdm <- glm(y ~ x1 + x2 + x3, data = train, family = binomial)

ggpubr::ggarrange(
  plotROC(sdm, test),
  plotImportance(sdm, test, n.perm = 20),
  labels = c("A", "B"), widths = c(1, 1.2)
)
```

A ROC curve says how well the model *ranks*; it does not tell you where to cut.
`plotThreshold()` does — sensitivity and specificity against the cutoff, with
the TSS-maximising threshold marked:

```{r fancyfx-threshold, fig.height = 4}
plotThreshold(sdm, test)
```

And neither says whether the probabilities themselves are *honest*.
`plotCalibration()` does: a model that says 0.7 should be right about 70% of
the time.

```{r fancyfx-calibration, fig.height = 5}
plotCalibration(sdm, test)
```

This is a genuinely separate question from discrimination, and AUC cannot
answer it. AUC only cares about ranking, so it is unchanged by any monotone
rescaling of the predictions — a model can post an excellent AUC while every
probability it reports is far too extreme. If those probabilities feed a
decision, an area calculation, or an expected count, calibration is the
property that matters. The reported slope makes it concrete: 1 is perfect,
below 1 means over-confident.

Note the rug here too. Calibration is usually worst at the extremes, and the
extremes usually hold the fewest predictions — so the most eye-catching
departures from the diagonal are often the least trustworthy points on the
plot. The rug and the interval on each bin both say so.

`calc_deviance()` covers what AUC cannot: AUC only cares about ranking and will
not notice predictions that are ordered correctly but wrong. Deviance penalises
confident mistakes hardest, and against the null deviance gives the proportion
explained that boosted-regression-tree work reports as standard.

### Is the hold-out actually independent?

`spatial_sorting_bias()` checks the assumption every metric above rests on. It
is the ratio of two mean nearest-neighbour distances — test presences to
training presences, over test absences to training presences.

```r
spatial_sorting_bias(test.presences, test.absences, training.presences)
#>  presence   absence       ssb
#>     0.381    20.434     0.019
```

Near 1, the split is doing its job. Near 0, the test presences sit much closer
to the training data than the absences do, and a model can score well on
proximity alone — its AUC is measuring the split rather than the species. This
is the quantity behind the cross-validation warnings above: they say the problem
exists, this says how bad it is for your split.

### Evaluation data is required, on purpose

`newdata` is a required argument, and passing the training data warns and
annotates the figure as in-sample. This is deliberate: an in-sample ROC can look
excellent for a model with no predictive value at all, and the easiest figure to
produce should not be the misleading one.

Cross-validation folds are supported via `folds`, drawn per fold so the spread
is visible rather than averaged away — but they come with a note, because
cross-validated metrics are weaker evidence than an independent hold-out. For
spatial models the gap is wider still: with random folds a held-out point
usually has a near neighbour among the training folds, so the model has in
effect already seen it. Use spatially blocked folds.

```r
plotROC(sdm, test, folds = spatial_block_ids)
```

Two caveats the functions state in their own documentation, worth repeating:
AUC and TSS are defined for **binary outcomes only** and a continuous response
is refused rather than scored; and permutation importance **splits credit badly
between correlated predictors**, which bites hard on environmental covariates.

## Spatial projections

A projection map is a persuasive object. It fills the study area with colour,
looks identical whether the model had a thousand observations in a region or
none, and nothing on it separates the part built on evidence from the part built
on the model's willingness to keep predicting.

Two maps put that distinction back. `terra` is a suggested package, so nothing
here is installed for users who never project.

```{r fancyfx-spatial, eval = requireNamespace("terra", quietly = TRUE), fig.height = 3.6}
library(terra)

set.seed(1)
grid <- rast(nrows = 50, ncols = 65, xmin = -71, xmax = -65,
             ymin = 41, ymax = 45, crs = "EPSG:4326")
lon <- init(grid, "x"); lat <- init(grid, "y")

sst <- 14 - 1.2 * (lat - 41) + 0.3 * (lon + 71); names(sst) <- "sst"
depth <- 20 + 30 * (lon + 71) + 25 * (45 - lat); names(depth) <- "depth"
covariates <- c(sst, depth)

# A survey covering only the north-west of the domain
points <- data.frame(lon = runif(400, -71, -67.5), lat = runif(400, 42, 45))
survey <- cbind(points, extract(covariates, points, ID = FALSE))
survey$present <- rbinom(400, 1,
                         plogis(-4 + 0.45 * survey$sst - 0.01 * survey$depth))

# A bootstrap ensemble of projections
ensemble <- rast(lapply(1:6, function(i) {
  refit <- mgcv::gam(present ~ s(sst) + s(depth),
                     data = survey[sample(nrow(survey), replace = TRUE), ],
                     family = binomial)
  terra::predict(covariates, refit, type = "response")
}))

ggpubr::ggarrange(
  plotUncertainty(ensemble, title = "Ensemble disagreement"),
  plotExtrapolation(covariates, survey, title = "Novel conditions"),
  labels = c("A", "B")
)
```

The south-east is both where the ensemble disagrees most **and** where the
projection has left the surveyed envelope — the honest reading being that the
model has nothing to say about it.

`plotExtrapolation()` draws a MESS surface: below zero, a cell is outside the
training range of at least one covariate. It works one covariate at a time,
though, so it cannot see novel *combinations* of individually ordinary values —
treat a clean surface as the absence of one specific problem, not permission to
project.

Also here: `hex_bin()` and `plotHexbin()` aggregate a raster or point data into
a hexagonal lattice; `thin_points()` thins records where clustering reflects
survey effort rather than the species; and `niche_overlap()` with
`niche_equivalency()` compare two predicted distributions against a
randomisation null. See `vignette("spatial")`.

## Supported models

| Model | Backend | What you get |
|---|---|---|
| `mgcv::gam()`, `bam()` | `gratia` | Partial effect, link scale, centered |
| `gamm4::gamm4()`, `mgcv::gamm()` | `gratia` | Partial effect, population level |
| `scam::scam()` | `gratia` | Partial effect, shape constraint preserved |
| `lm()`, `glm()` | `marginaleffects` | Predicted values |
| `lme4::lmer()`, `glmer()` | `marginaleffects` | Predicted values, population level |
| `glmmTMB::glmmTMB()` | `marginaleffects` | Predicted values, population level |
| `brms::brm()` | `marginaleffects` | Predicted values, credible interval |
| `rstanarm::stan_glm()`, `stan_glmer()` | `marginaleffects` | Predicted values, credible interval |
| Most other fitted models | `marginaleffects` | Predicted values |

Support beyond those comes from whatever `marginaleffects` handles. The rows
above are the families verified against real fits.

Everything in the GAM family reports a **partial effect** by default, so those
rows are comparable with each other. Two get there only because the package
intervenes: `scam` inherits from `glm` rather than `gam` and would otherwise
fall through to predictions, and `gamm4`/`gamm` return a *list* holding a GAM
rather than a fitted model — `formula()`, `predict()` and `marginaleffects` all
refuse the wrapper, so it is unwrapped to its `$gam` first. The random effects
stay behind with the wrapper, which is why those rows say population level.

**Bayesian fits** are summarised from posterior draws, which give an interval
but no standard error — so the ribbon is the credible interval at `level`, and
there is no `±1 SE` ribbon to be had. Write `interval = "cri"` if you would
rather say so explicitly; it computes the same thing. `brms` takes `re_formula`
rather than `re.form`, and that translation is handled for you.

For a **mixed model**, `re.form` defaults to `NA`, so the effect is drawn at the
population level. This matters: left to the backend's own default, the grouping
factor is held at its modal level and the plot silently shows the effect for one
arbitrary group. Note that the ribbon covers uncertainty in the fixed effects
only.

### A caveat worth reading

The two backends compute **different quantities**. A partial effect is one
term's contribution in isolation, centered so it averages to zero. A predicted
value is the model's fitted output with the other predictors held at
representative values. `fancyfx` labels the y axis with whichever it computed,
and you should not compare them as though they were on the same footing.

The default `interval = "se"` ribbon spans roughly 68%, not 95% — it is this
package's historical default, kept so existing figures do not silently change.
Pass `interval = "ci"` for a conventional confidence interval.

See `vignette("fancyfx")` for the full discussion.

## Migrating from plotSmooths()

```r
# Old:
plotSmooths(gam.fit, iris, "Sepal.Length")

# New — same arguments, same output:
plotEffects(gam.fit, iris, "Sepal.Length")
```

The defaults reproduce exactly what `plotSmooths()` drew for a GAM, so
migrating is a rename and nothing more.

## How to cite

```r
citation("fancyfx")
```

`fancyfx` is a plotting layer over other people's estimation work. Please also
cite whichever package computed your effects: `gratia` and `mgcv` for GAM
partial effects, `marginaleffects` otherwise.
