---
title: "Evaluating a model"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Evaluating a model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup}
library(fancyfx)
```

An effect plot says what a model claims. This vignette is about the other
question: whether to believe it.

Everything here is aimed at **presence/absence models** — the species
distribution case, where the response is whether something was there. AUC and
TSS are defined for a binary outcome and nothing else, so applied to a Gaussian
model of biomass they would return a number with no meaning. `fancyfx` refuses
that rather than computing it.

## A worked example

```{r data}
set.seed(1)
d <- data.frame(x1 = runif(600, 1, 10),
                x2 = runif(600, 1, 10),
                x3 = runif(600, 1, 10))
# Only x1 carries any signal.
d$y <- rbinom(600, 1, plogis(-3 + 0.6 * d$x1))

train <- d[1:300, ]
test  <- d[301:600, ]

sdm <- glm(y ~ x1 + x2 + x3, data = train, family = binomial)
```

## The most important thing on this page

`newdata` is a **required** argument to every function here. There is no default,
and the default is not the training data.

That is deliberate. A model scored against the data it was fitted to flatters
itself, sometimes enormously, and an in-sample ROC curve can look excellent for
a model with no predictive value at all. Making that the easiest figure to
produce would be a disservice.

Passing the training data still works — refusing outright would be obstructive,
and sometimes it is genuinely what you want — but it warns, and the figure is
annotated so it cannot be quietly published as validation:

```{r in-sample, fig.height = 5, warning = FALSE}
plotROC(sdm, train, title = "Scored on the training data")
```

Read the caption. Compare it with the honest version:

```{r out-of-sample, fig.height = 5}
plotROC(sdm, test, title = "Scored on held-out data")
```

## Discrimination: the ROC curve

The ROC curve traces sensitivity against the false positive rate as the decision
threshold moves. The diagonal is chance. AUC summarises the whole curve into the
probability that a randomly chosen presence is ranked above a randomly chosen
absence.

`fancyfx` computes AUC from ranks — the Mann-Whitney form — rather than by
integrating the drawn curve, so tied predictions are handled exactly. The curve
itself is drawn as a step function, because that is what an empirical ROC is;
joining the corners would draw operating points the model cannot reach.

The numbers are available on their own:

```{r metrics}
m <- threshold_metrics(sdm, test)

attr(m, "auc")
attr(m, "prevalence")
head(m, 3)
```

## Choosing a cutoff

AUC is threshold-free, which is its strength and its limitation: it says how
well the model ranks, and nothing about where to cut. For that:

```{r threshold}
plotThreshold(sdm, test)
```

Sensitivity and specificity trade off against each other, and TSS
[@allouche2006] is their sum less one — so its peak, marked with the dashed
line, is the cutoff balancing them best. TSS is preferred to raw accuracy in
species distribution work because it is not inflated by a rare positive class.

That peak is *one* defensible choice, not the only one. If a false absence costs
more than a false presence — a missed occurrence of something you are trying to
protect — then the right cutoff is not where TSS peaks. Both curves are drawn so
that judgement can be made rather than assumed.

```{r threshold-tss}
# TSS alone, without the curves it is built from
plotThreshold(sdm, test, metrics = "tss", mark.best = TRUE)
```

## Cross-validation, and why it is weaker

Folds are supported, and drawn per fold so the spread is visible rather than
averaged into a single reassuring number:

```{r folds, fig.height = 5}
set.seed(2)
folds <- sample(rep(1:4, length.out = nrow(test)))

plotROC(sdm, test, folds = folds)
```

The first time you use `folds` in a session, the package says something you
should read. Cross-validated metrics are **weaker evidence than an independent
hold-out**: the folds come from the same sample, and the model saw the rest of
it. A cross-validated score speaks to how stable the fit is, not to how it will
behave somewhere new.

For spatial models the gap is wider still. With random folds, a held-out point
usually has a near neighbour among the training folds, so the model has in
effect already seen it, and the score is optimistic for reasons that have
nothing to do with the model being good. **Spatially blocked folds** are the
honest version — blocks of contiguous space held out together, so the
distance between training and testing data is real. Packages such as
`blockCV` generate them; pass the resulting fold IDs straight in.

The spread across folds is the part worth reading. If the AUC range is wide, or
the TSS-maximising cutoff moves substantially between folds, then reporting a
single threshold to three decimal places is false precision.

## Variable importance

Which predictors is the model actually leaning on?

```{r importance, fig.height = 3.5}
plotImportance(sdm, test, n.perm = 20)
```

The measure is **permutation importance** [@breiman2001]: shuffle one predictor,
breaking its relationship with the response while leaving its distribution
intact, and see how much worse the model scores. For a binary response the
metric is the drop in AUC; otherwise the increase in RMSE. Either way, bigger
means the model was relying on it more.

It is model-agnostic — it needs nothing but the ability to predict — so a GAM, a
GLM, a mixed model and a Bayesian fit all give numbers that mean the same thing
and can be compared.

The point is drawn at the mean and the line spans the permutations. The **spread
is the point**: a variable whose range crosses the dashed zero line did no
measurable work, and a bar chart of means would hide that. Here `x1` is doing
everything and `x2` and `x3` are doing nothing, which is exactly how the data
were simulated.

Because permutation is random, results are seeded by default — an unseeded
figure cannot be redrawn. The caller's random stream is restored afterwards, so
this does not silently change results elsewhere in your script.

### Two things it does not tell you

**Correlated predictors share credit badly.** If two variables carry much the
same information, permuting either one alone barely hurts, because the other
still carries it. Both look unimportant and the pair is not. This bites hard on
environmental covariates, which are routinely collinear — sea surface
temperature and depth, say. If a variable you expect to matter shows near-zero
importance, check what it is correlated with before concluding anything.

**It measures use, not effect size.** A variable can be important here and have
an effect too small to care about, or the reverse. Read it beside
`plotEffects()`, not instead of it:

```{r importance-and-effect, fig.height = 4}
ggpubr::ggarrange(
  plotImportance(sdm, test, n.perm = 20),
  plotEffects(sdm, train, "x1", scale = "response",
              ylab = "P(presence)"),
  labels = c("A", "B"), widths = c(1, 1.1)
)
```

## Putting it in a paper

Everything here uses the same theme, palette and panel machinery as the effect
plots, so an evaluation figure and an effect figure sit together without
adjustment. See the "Publication-ready output" section of `vignette("fancyfx")`
for `base_size`, panel labels and saving at a fixed size.

## References
