---
title: "Evaluating a model"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Evaluating a model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup}
library(fancyfx)
```

An effect plot says what a model claims. This vignette is about the other
question: whether to believe it.

Most of what follows is aimed at **presence/absence models** — the species
distribution case, where the response is whether something was there. AUC, TSS
and calibration are defined for a binary outcome and nothing else, so applied to
a Gaussian model of biomass they would return a number with no meaning.
`fancyfx` refuses that rather than computing it. Permutation importance is the
exception: it works for any model.

There are three distinct questions here, and it is worth keeping them apart:

| Question | Function |
|---|---|
| Does it rank presences above absences? | `plotROC()` |
| Where should the cutoff go? | `plotThreshold()` |
| Are the probabilities honest? | `plotCalibration()` |
| Which predictors is it using? | `plotImportance()` |

The first three are not interchangeable. A model can rank perfectly and still
report probabilities that are wildly wrong.

## A worked example

```{r data}
set.seed(1)
d <- data.frame(x1 = runif(600, 1, 10),
                x2 = runif(600, 1, 10),
                x3 = runif(600, 1, 10))
# Only x1 carries any signal.
d$y <- rbinom(600, 1, plogis(-3 + 0.6 * d$x1))

train <- d[1:300, ]
test  <- d[301:600, ]

sdm <- glm(y ~ x1 + x2 + x3, data = train, family = binomial)
```

## The most important thing on this page

`newdata` is a **required** argument to every function here. There is no default,
and the default is not the training data.

That is deliberate. A model scored against the data it was fitted to flatters
itself, sometimes enormously, and an in-sample ROC curve can look excellent for
a model with no predictive value at all. Making that the easiest figure to
produce would be a disservice.

Passing the training data still works — refusing outright would be obstructive,
and sometimes it is genuinely what you want — but it warns, and the figure is
annotated so it cannot be quietly published as validation:

```{r in-sample, fig.height = 5, warning = FALSE}
plotROC(sdm, train, title = "Scored on the training data")
```

Read the caption. Compare it with the honest version:

```{r out-of-sample, fig.height = 5}
plotROC(sdm, test, title = "Scored on held-out data")
```

## Discrimination: the ROC curve

The ROC curve traces sensitivity against the false positive rate as the decision
threshold moves. The diagonal is chance. AUC summarises the whole curve into the
probability that a randomly chosen presence is ranked above a randomly chosen
absence.

`fancyfx` computes AUC from ranks — the Mann-Whitney form — rather than by
integrating the drawn curve, so tied predictions are handled exactly. The curve
itself is drawn as a step function, because that is what an empirical ROC is;
joining the corners would draw operating points the model cannot reach.

The numbers are available on their own:

```{r metrics}
m <- threshold_metrics(sdm, test)

attr(m, "auc")
attr(m, "prevalence")
head(m, 3)
```

## Choosing a cutoff

AUC is threshold-free, which is its strength and its limitation: it says how
well the model ranks, and nothing about where to cut. For that:

```{r threshold}
plotThreshold(sdm, test)
```

Sensitivity and specificity trade off against each other, and TSS
[@allouche2006] is their sum less one — so its peak, marked with the dashed
line, is the cutoff balancing them best. TSS is preferred to raw accuracy in
species distribution work because it is not inflated by a rare positive class.

That peak is *one* defensible choice, not the only one. If a false absence costs
more than a false presence — a missed occurrence of something you are trying to
protect — then the right cutoff is not where TSS peaks. Both curves are drawn so
that judgement can be made rather than assumed.

```{r threshold-tss}
# TSS alone, without the curves it is built from
plotThreshold(sdm, test, metrics = "tss", mark.best = TRUE)
```

## Calibration: are the probabilities honest?

Discrimination and calibration are different properties, and a model can be
good at one and bad at the other.

AUC only cares whether presences are ranked above absences. That makes it blind
to any monotone rescaling of the predictions: a model can post an excellent AUC
while every probability it reports is far too extreme. If those probabilities
are only ever compared with each other, that may not matter. If they feed a
decision, an area of suitable habitat, or an expected count, it matters a great
deal.

```{r calibration, fig.height = 5}
plotCalibration(sdm, test)
```

A well calibrated model sits on the diagonal: when it says 0.7, the outcome
happens about 70% of the time. The reported **slope** makes that concrete — 1
is perfect, and below 1 means the predictions are too extreme.

To see what a failure looks like, here is the same model with its coefficients
inflated. The ranking is untouched, so its AUC is *identical*:

```{r calibration-bad, fig.height = 5}
overconfident <- sdm
overconfident$coefficients <- overconfident$coefficients * 2.5

c(original = attr(threshold_metrics(sdm, test), "auc"),
  overconfident = attr(threshold_metrics(overconfident, test), "auc"))

plotCalibration(overconfident, test, title = "Same AUC, useless probabilities")
```

The curve is flatter than the diagonal and the rug has piled up against 0 and
1. AUC saw none of this.

### Reading the plot honestly

Two features of the plot exist to stop it being over-read:

* **The rug.** Calibration is usually worst at the extremes, and the extremes
  usually hold the fewest predictions — so the most eye-catching departures
  from the diagonal are often the least trustworthy points. The rug shows where
  the predictions actually are.
* **The interval on each bin**, which is a Wilson score interval rather than a
  normal approximation. With few observations in a bin, or an observed
  frequency near 0 or 1 — both routine here — the normal approximation puts the
  bounds outside `[0, 1]`, and a plot should not claim something impossible.

Binning is a genuine trade-off, exposed as `binning`. Quantile bins (the
default) hold equal numbers of observations, so every point is equally
reliable, at the cost of varying bin widths. Equal-width bins are easier to
read but leave the extremes nearly empty — precisely where the interesting
behaviour is.

```{r calibration-binning, fig.height = 5}
plotCalibration(sdm, test, binning = "width", bins = 8)
```

The numbers are available on their own, including the Brier score — the mean
squared error of the probabilities, covering calibration and discrimination in
one figure:

```{r calibration-numbers}
cal <- calibration_estimates(sdm, test)

attr(cal, "calibration")
attr(cal, "brier")
head(cal, 3)
```

## Cross-validation, and why it is weaker

Folds are supported, and drawn per fold so the spread is visible rather than
averaged into a single reassuring number:

```{r folds, fig.height = 5}
set.seed(2)
folds <- sample(rep(1:4, length.out = nrow(test)))

plotROC(sdm, test, folds = folds)
```

The first time you use `folds` in a session, the package says something you
should read. Cross-validated metrics are **weaker evidence than an independent
hold-out**: the folds come from the same sample, and the model saw the rest of
it. A cross-validated score speaks to how stable the fit is, not to how it will
behave somewhere new.

For spatial models the gap is wider still. With random folds, a held-out point
usually has a near neighbour among the training folds, so the model has in
effect already seen it, and the score is optimistic for reasons that have
nothing to do with the model being good. **Spatially blocked folds** are the
honest version — blocks of contiguous space held out together, so the
distance between training and testing data is real. Packages such as
`blockCV` generate them; pass the resulting fold IDs straight in.

The spread across folds is the part worth reading. If the AUC range is wide, or
the TSS-maximising cutoff moves substantially between folds, then reporting a
single threshold to three decimal places is false precision.

## Deviance

AUC only cares about ranking, so it cannot see predictions that are ordered
correctly and still wrong. Deviance can, and it penalises confident mistakes
hardest — usually the failure that matters.

```{r deviance}
predicted <- predict(sdm, test, type = "response")

fitted.deviance <- calc_deviance(test$y, predicted, "binomial")
null.deviance <- calc_deviance(test$y, rep(mean(test$y), nrow(test)),
                               "binomial")

# Proportion of deviance explained
1 - fitted.deviance / null.deviance
```

It is most useful as a relative measure. Against the null deviance — what you
would get predicting the overall mean for everything — the ratio is the
proportion of deviance explained, the number boosted-regression-tree work
reports as a matter of course. `family` also takes `"poisson"`, `"gaussian"`
and `"laplace"`, so it is not restricted to presence/absence.

## How independent is the hold-out, really?

Everything above assumes the evaluation data tells you something the training
data did not. `spatial_sorting_bias()` checks that assumption instead of taking
it on faith.

The statistic is the ratio of two mean nearest-neighbour distances: from each
test presence to the closest training presence, over the same for each test
absence.

```{r ssb}
set.seed(1)
training <- cbind(runif(150, 0, 10), runif(150, 0, 10))

# A split whose presences sit on top of the training data
near <- cbind(runif(60, 0, 10), runif(60, 0, 10))
background <- cbind(runif(60, 0, 40), runif(60, 0, 40))

spatial_sorting_bias(near, background, training)
```

Near 1 means the test presences and test absences are equally far from the
training data — the split is doing its job. Near 0 means the presences are much
closer, and a model can score well by knowing roughly where the training data
was without knowing anything about the species. Its AUC is then measuring the
split rather than the species.

This is what the cross-validation warnings elsewhere in the package are about.
They say the problem exists; this says how bad it is for a particular split,
which is the version worth putting in a methods section. Pass `geo = TRUE` for
longitude and latitude, and distances become great-circle.

A low value is not a reason to throw the model away. It is a reason to use
spatially blocked folds, or to sample the test absences to match the presences'
distance distribution — and then to say which you did.

## Variable importance

Which predictors is the model actually leaning on?

```{r importance, fig.height = 3.5}
plotImportance(sdm, test, n.perm = 20)
```

The measure is **permutation importance** [@breiman2001]: shuffle one predictor,
breaking its relationship with the response while leaving its distribution
intact, and see how much worse the model scores. For a binary response the
metric is the drop in AUC; otherwise the increase in RMSE. Either way, bigger
means the model was relying on it more.

It is model-agnostic — it needs nothing but the ability to predict — so a GAM, a
GLM, a mixed model and a Bayesian fit all give numbers that mean the same thing
and can be compared.

`permutation_importance()` gives the numbers behind the plot — one row per
variable per permutation, so you can summarise them your own way:

```{r importance-numbers}
imp <- permutation_importance(sdm, test, n.perm = 20)

aggregate(.importance ~ .variable, data = imp, FUN = mean)
attr(imp, "metric")
```

The point is drawn at the mean and the line spans the permutations. The **spread
is the point**: a variable whose range crosses the dashed zero line did no
measurable work, and a bar chart of means would hide that. Here `x1` is doing
everything and `x2` and `x3` are doing nothing, which is exactly how the data
were simulated.

Because permutation is random, results are seeded by default — an unseeded
figure cannot be redrawn. The caller's random stream is restored afterwards, so
this does not silently change results elsewhere in your script.

### Two things it does not tell you

**Correlated predictors share credit badly.** If two variables carry much the
same information, permuting either one alone barely hurts, because the other
still carries it. Both look unimportant and the pair is not. This bites hard on
environmental covariates, which are routinely collinear — sea surface
temperature and depth, say. If a variable you expect to matter shows near-zero
importance, check what it is correlated with before concluding anything.

**It measures use, not effect size.** A variable can be important here and have
an effect too small to care about, or the reverse. Read it beside
`plotEffects()`, not instead of it:

```{r importance-and-effect, fig.height = 4}
ggpubr::ggarrange(
  plotImportance(sdm, test, n.perm = 20),
  plotEffects(sdm, train, "x1", scale = "response",
              ylab = "P(presence)"),
  labels = c("A", "B"), widths = c(1, 1.1)
)
```

## Putting it in a paper

Everything here uses the same theme, palette and panel machinery as the effect
plots, so an evaluation figure and an effect figure sit together without
adjustment. See the "Publication-ready output" section of `vignette("fancyfx")`
for `base_size`, panel labels and saving at a fixed size.

## References
