---
title: "Getting started with fancyfx"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Getting started with fancyfx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup}
library(fancyfx)
```

## The idea

An effect curve on its own is easy to over-read. A model will happily draw a
confident-looking bend at the far right of the x axis, and nothing in the plot
tells you that only three observations sit under it.

`fancyfx` pairs every effect curve with a rug of the raw data, drawn directly
above it on a shared x axis. The shape of the effect and the weight of evidence
behind it end up in the same figure, so they get read together.

The second thing the package is for is getting that figure into a manuscript
without a further round of fiddling. Defaults are chosen for publication rather
than exploration — see [Publication-ready output](#publication-ready-output) —
and each of them is an argument, so a house style can replace any of them.

```{r idea}
gam.fit <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)

plotEffects(gam.fit, iris, "Sepal.Length",
            xlab = "Sepal length (cm)")
```

The histogram along the top is the whole point. Where it is thin, be careful.

## GAMs

Generalized additive models fitted with **mgcv** [@wood2017] are the package's
original use case, and remain the default path. Effects are extracted with
`gratia::smooth_estimates()` [@gratia], which reports the *partial effect* of a
smooth: the term's own contribution to the linear predictor, centered so it
averages to zero.

`combinePlots()` runs the same thing across several terms and arranges them as
labelled panels. Each panel keeps its own rug, since the predictors do not
generally share a distribution.

```{r gam-multi}
gam.fit2 <- mgcv::gam(Petal.Length ~ s(Sepal.Length) + s(Petal.Width),
                      data = iris)

combinePlots(gam.fit2, iris,
             vars = c("Sepal.Length", "Petal.Width"),
             title = "Partial effects on petal length")
```

## The rest of the GAM family

`bam()` for large data, `gamm4()` and `gamm()` for GAMs fitted through a mixed
model, and `scam()` for shape-constrained smooths all report a **partial
effect**, so they are comparable with each other and with `gam()`:

```{r gam-family, eval = requireNamespace("scam", quietly = TRUE), fig.height = 4}
big <- mgcv::bam(Petal.Length ~ s(Sepal.Length), data = iris)

# A monotone-increasing smooth: the constraint is part of the fit,
# and survives into the plot.
constrained <- scam::scam(Petal.Length ~ s(Sepal.Length, bs = "mpi"),
                          data = iris)

comparePlots(list("Unconstrained (bam)" = big,
                  "Monotone (scam)" = constrained),
             iris, "Sepal.Length")
```

Two of these need the package to intervene, which is worth knowing if you ever
hit the same classes elsewhere:

* **`scam`** objects inherit from `glm`, not `gam`. Without a method of their
  own they would fall through to the prediction backend and quietly report a
  different quantity than every other GAM here.
* **`gamm4()` and `gamm()`** do not return a fitted model at all. They return a
  list holding the GAM alongside the mixed-model fit it was estimated through.
  `formula()`, `predict()` and `marginaleffects` all refuse that wrapper, so
  `fancyfx` unwraps it to its `$gam` element first.

The random effects stay behind with the wrapper, so a `gamm4` smooth is drawn
at the population level — consistent with how the package treats mixed models
generally, and the same caveat applies: the ribbon covers uncertainty in the
smooth, not variation between groups.

## Other models

Anything that is not an mgcv GAM is routed through
`marginaleffects::predictions()` [@marginaleffects], which supports a wide range
of model classes. Nothing about the call changes:

```{r lm}
lm.fit <- lm(mpg ~ wt + hp, data = mtcars)

combinePlots(lm.fit, mtcars, vars = c("wt", "hp"),
             rug.type = "density")
```

For a GLM, `scale = "response"` puts the curve back on the scale of the outcome
— probabilities, for a logistic regression — rather than log-odds:

```{r glm}
glm.fit <- glm(am ~ wt + hp, data = mtcars, family = binomial)

plotEffects(glm.fit, mtcars, "wt",
            scale = "response", interval = "ci",
            xlab = "Weight (1000 lbs)",
            ylab = "P(manual transmission)")
```

## Factor-smooth interactions

A factor-smooth interaction, `s(x, by = f)`, is not one smooth but one per level
of `f`. They are drawn as separate curves with a legend:

```{r factor-smooth}
by.species <- mgcv::gam(Petal.Length ~ s(Sepal.Length, by = Species) + Species,
                        data = iris)

plotEffects(by.species, iris, "Sepal.Length",
            xlab = "Sepal length (cm)")
```

Note the rug is shared: it shows the distribution of `Sepal.Length` overall, not
per species. A level with few observations will have a wide ribbon, which is the
signal to look for.

## Comparing models

`combinePlots()` holds the model fixed and varies the predictor.
`comparePlots()` does the opposite, which is how you ask whether a modelling
choice was worth making:

```{r compare, fig.height = 4}
plain <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)

comparePlots(list("Single smooth" = plain,
                  "Smooth by species" = by.species),
             iris, "Sepal.Length",
             title = "Is a factor-smooth interaction worth it?")
```

Panels are titled from the list's names, and labelled `A`, `B`, `C`. Models need
not share a class — but see the caveat below about comparing a partial effect
against a prediction.

## Mixed models

`lme4` and `glmmTMB` fits go through the same backend as any other non-GAM
model. The one thing worth knowing is what happens to the random effects:

```{r mixed, eval = requireNamespace("lme4", quietly = TRUE)}
set.seed(1)
d <- data.frame(g = rep(letters[1:8], each = 25), x = runif(200, 1, 10))
d$y <- 2 * d$x + rep(rnorm(8, 0, 5), each = 25) + rnorm(200)

fit <- lme4::lmer(y ~ x + (1 | g), data = d)

plotEffects(fit, d, "x")
```

`re.form` defaults to `NA`, so the curve is the **population-level** effect.
This is not a cosmetic default. Left to the backend's own, the grouping factor
is held at its modal level, and the plot silently shows the effect for whichever
group happens to be most common — a different curve entirely. Pass
`re.form = NULL` to include the random effects.

One limitation to state plainly: the ribbon reflects uncertainty in the fixed
effects only. It does not widen to account for variation between groups, so it
is narrower than a genuine prediction interval for a new group.

## Bayesian models

`brms` and `rstanarm` fits go through the same backend, and the call does not
change:

```r
fit <- brms::brm(y ~ x + (1 | g), data = d)

plotEffects(fit, d, "x")
```

Three things are different underneath, all handled for you:

* A posterior summary gives an **interval but no standard error**, so the ribbon
  is the credible interval at `level`. There is no `±1 SE` ribbon to be had, and
  asking for one says so rather than failing. `interval = "cri"` is accepted as
  a name for the same computation, if you would rather be explicit in your code.
* `brms` names its random-effects argument **`re_formula`**, not `re.form`.
  Write either; the translation happens for you.
* The bounded interval construction used for GLMs is **not available**, so a
  response-scale interval comes straight from the draws — which respects the
  bounds anyway, because the draws do.

As with any mixed model, `re.form` defaults to `NA` for the population-level
effect.

## What the y axis means

This is the section worth reading before you put a `fancyfx` figure in a paper.

The two backends compute **different quantities**, and the package does not
paper over the difference:

| Model | `scale = "auto"` gives you | Centered on zero? | Other predictors |
|---|---|---|---|
| mgcv GAM | Partial effect, link scale | Yes | Excluded |
| Everything else | Predicted value, response scale | No | Held at representative values |

A **partial effect** is one term's contribution in isolation. It is centered by
construction, so its zero line is an arbitrary reference point, not a
meaningful value of the outcome. A **predicted value** is the model's actual
fitted output as one predictor varies while the others sit at their means (for
numeric predictors) or modes (for factors).

Because they are different quantities, `fancyfx` labels the y axis with whichever
one it computed, and you should not compare them side by side as though they
were on the same footing.

A GAM asked for `scale = "response"` gets *predictions*, not a back-transformed
partial effect. That is deliberate: a centered partial effect has no coherent
back-transformation on its own, and exponentiating one would produce a curve
that looks meaningful and is not.

```{r quantities}
partial <- effect_estimates(gam.fit, "Sepal.Length")
predicted <- effect_estimates(gam.fit, "Sepal.Length", scale = "response")

c(partial = attr(partial, "quantity"),
  predicted = attr(predicted, "quantity"))

# The partial effect averages to zero; the prediction does not.
round(c(partial = mean(partial$.estimate),
        predicted = mean(predicted$.estimate)), 3)
```

### The ribbon

The default `interval = "se"` draws a `± 1` standard error band, which spans
roughly 68%. That was this package's default until version 0.10.0, which is
half the width of what `mgcv::plot.gam()` and `gratia::draw()` show and a width
most readers would assume was 95%. The default is now a 95% pointwise interval;
`interval = "se"` gives the narrow band, explicitly:

```{r intervals}
plotEffects(lm.fit, mtcars, "wt", interval = "ci", level = 0.95)
```

For a GAM smooth there is a third option, and it is the statistically careful
one. A pointwise interval covers the true value at each x separately. Across a
curve evaluated at a hundred points, the true function strays outside a
pointwise 95% band far more often than 5% of the time — so a pointwise ribbon
does not support a claim about the *shape* of the smooth, which is usually the
reason for drawing it.

```{r simultaneous, fig.height = 4}
gam.fit3 <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)

ggpubr::ggarrange(
  plotEffects(gam.fit3, iris, "Sepal.Length", interval = "ci",
              xlab = "Sepal length (cm)"),
  plotEffects(gam.fit3, iris, "Sepal.Length", interval = "simultaneous",
              xlab = "Sepal length (cm)"),
  labels = c("Pointwise", "Simultaneous"),
  font.label = list(size = 11)
)
```

The simultaneous band is noticeably wider, which is the point. It is available
only for GAM partial effects, where `gratia` can simulate from the posterior of
the smooth; asking for one anywhere else is an error rather than a silent
substitution. Because it is simulated, it is seeded so a figure can be redrawn,
and the seed is restored afterwards so nothing else in your script shifts.

On the response scale, `interval = "ci"` builds the interval on the link scale
and back-transforms it, so it stays inside the range the response actually
admits — a probability band will not run past 0 or 1. That construction is
asymmetric about the estimate and so has no single standard error behind it,
which is why `interval = "se"` cannot use it and may produce a band that does
escape those bounds. If you are plotting probabilities, prefer `"ci"`.

Under a Bayesian fit, `marginaleffects` returns a credible interval, and
`interval = "ci"` will give you that rather than a confidence interval — the
argument names the ribbon, not the inferential philosophy.

## Transforms and rugs

`transform` is applied to both the curve and the rug, so the two stay on the
same x axis. Applying it to only one of them would defeat the entire purpose of
stacking them.

```{r transform}
plotEffects(lm.fit, mtcars, "hp",
            transform = "log10",
            xlab = "Horsepower (log10)")
```

Rugs come in two styles. `"histogram"` (the default) shows counts and reads well
for moderate sample sizes; `"density"` is smoother and works better when the
data are dense enough that a histogram becomes noisy. Side by side:

```{r rug-types, fig.height = 4}
ggpubr::ggarrange(
  plotEffects(lm.fit, mtcars, "wt", rug.type = "histogram",
              xlab = "Weight (1000 lbs)"),
  plotEffects(lm.fit, mtcars, "wt", rug.type = "density",
              xlab = "Weight (1000 lbs)"),
  labels = c("A", "B")
)
```

`plotRugs()` will also give you the rug on its own, if you would rather compose
the figure yourself:

```{r rug-alone, fig.height = 1.5}
plotRugs(mtcars, "wt", type = "density")
```

`combinePlots()` accepts one transform for all variables or one per variable:

```{r per-var-transform}
combinePlots(lm.fit, mtcars, vars = c("wt", "hp"),
             var.transform = c("none", "log10"))
```

## Publication-ready output

The default theme is `theme_fancyfx()`, built on `ggpubr::theme_pubr()`: no
background panel, no grid, plain black axis lines, and text sized to survive
being shrunk into a journal column.

The main thing to adjust is `base_size`, and the reason is not obvious: a figure
saved narrow gets scaled *up* on the page, so its text ends up smaller than the
same text in a wide figure. Set the size to match the width you will save at.

```{r theme-size, fig.height = 4}
plotEffects(lm.fit, mtcars, "wt",
            xlab = "Weight (1000 lbs)",
            theme = theme_fancyfx(base_size = 14))
```

`base_size` scales everything at once and keeps the relative sizes balanced,
which is usually what you want. When it is not, every element takes its own
argument:

```{r theme-elements, fig.height = 4}
plotEffects(lm.fit, mtcars, "wt",
            xlab = "Weight (1000 lbs)",
            theme = theme_fancyfx(base_size = 13,
                                  axis.title.size = 18,
                                  axis.text.size = 10,
                                  legend.title.size = 15))
```

The full set is `axis.title.size`, `axis.text.size`, `title.size`,
`subtitle.size`, `caption.size`, `legend.title.size`, `legend.text.size` and
`strip.text.size`. Anything left alone follows `base_size`.

Two pieces of text are drawn by the arranging step rather than the theme, so
they have their own arguments on `combinePlots()` and `comparePlots()` — the
panel labels and the overall figure title:

```{r theme-labels, fig.height = 4}
combinePlots(lm.fit, mtcars, vars = c("wt", "hp"),
             title = "Both terms",
             label.size = 20, title.size = 18,
             theme = theme_fancyfx(base_size = 14))
```

Any other ggplot2 theme can be passed instead:

```r
plotEffects(lm.fit, mtcars, "wt", theme = ggplot2::theme_minimal())
```

Panels are labelled `A`, `B`, `C` by default. Journals vary, so the style is an
argument:

```r
combinePlots(fit, dat, vars, labels = "a")                     # a, b, c
combinePlots(fit, dat, vars, labels = "1")                     # 1, 2, 3
combinePlots(fit, dat, vars, labels = "none")                  # no labels
combinePlots(fit, dat, vars, labels = c("Panel one", "Panel two"))
```

### The palette

Curves that split by a factor use `fancyfx_palette()`:

```{r palette}
fancyfx_palette()
```

It was chosen by search rather than by eye, against the properties that decide
whether a reader can actually tell two curves apart: every colour sits in a mid
lightness band, carries enough chroma not to read as grey, clears 3:1 contrast
against a white page, and stays separable under simulated protanopia and
deuteranopia. A legend is always drawn, so identity never rests on colour alone.

Six is the limit. Past that, colours stop being tellable apart however they are
chosen, and `plotEffects()` says so rather than inventing a seventh — a facet
per level communicates far better.

Saving at a fixed size, as a journal will ask for:

```r
ggplot2::ggsave("figure-1.pdf", p, width = 174, height = 100, units = "mm")
```

## Migrating from plotSmooths()

`plotSmooths()` was the GAM-only ancestor of `plotEffects()`, from when the
package handled nothing else. It still works and produces an identical plot, but
warns once per session:

```r
# Old:
plotSmooths(gam.fit, iris, "Sepal.Length")

# New — same arguments, same output:
plotEffects(gam.fit, iris, "Sepal.Length")
```

The defaults (`scale = "auto"`, `interval = "se"`) reproduce exactly what
`plotSmooths()` drew for a GAM, so migrating is a rename and nothing more.

## Evaluating the model

Everything above is about what a model *claims*. Whether it has earned the claim
is a separate question, with its own functions — `plotROC()`, `plotThreshold()`
and `plotImportance()` — and its own vignette:

```r
vignette("evaluation")
```

They share this package's theme, palette and panel machinery, so an evaluation
figure and an effect figure sit together in one manuscript without adjustment.

## How to cite

```r
citation("fancyfx")
```

`fancyfx` is a plotting layer over other people's estimation work. Please also
cite whichever package actually computed your effects: **gratia** and **mgcv**
for GAM partial effects, **marginaleffects** otherwise.

## References
