---
title: "Getting started with fancyfx"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Getting started with fancyfx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup}
library(fancyfx)
```

## The idea

An effect curve on its own is easy to over-read. A model will happily draw a
confident-looking bend at the far right of the x axis, and nothing in the plot
tells you that only three observations sit under it.

`fancyfx` pairs every effect curve with a rug of the raw data, drawn directly
above it on a shared x axis. The shape of the effect and the weight of evidence
behind it end up in the same figure, so they get read together.

```{r idea}
gam.fit <- mgcv::gam(Petal.Length ~ s(Sepal.Length), data = iris)

plotEffects(gam.fit, iris, "Sepal.Length",
            xlab = "Sepal length (cm)")
```

The histogram along the top is the whole point. Where it is thin, be careful.

## GAMs

Generalized additive models fitted with **mgcv** [@wood2017] are the package's
original use case, and remain the default path. Effects are extracted with
`gratia::smooth_estimates()` [@gratia], which reports the *partial effect* of a
smooth: the term's own contribution to the linear predictor, centered so it
averages to zero.

`combinePlots()` runs the same thing across several terms and arranges them as
labelled panels. Each panel keeps its own rug, since the predictors do not
generally share a distribution.

```{r gam-multi}
gam.fit2 <- mgcv::gam(Petal.Length ~ s(Sepal.Length) + s(Petal.Width),
                      data = iris)

combinePlots(gam.fit2, iris,
             vars = c("Sepal.Length", "Petal.Width"),
             title = "Partial effects on petal length")
```

## Other models

Anything that is not an mgcv GAM is routed through
`marginaleffects::predictions()` [@marginaleffects], which supports a wide range
of model classes. Nothing about the call changes:

```{r lm}
lm.fit <- lm(mpg ~ wt + hp, data = mtcars)

combinePlots(lm.fit, mtcars, vars = c("wt", "hp"),
             rug.type = "density")
```

For a GLM, `scale = "response"` puts the curve back on the scale of the outcome
— probabilities, for a logistic regression — rather than log-odds:

```{r glm}
glm.fit <- glm(am ~ wt + hp, data = mtcars, family = binomial)

plotEffects(glm.fit, mtcars, "wt",
            scale = "response", interval = "ci",
            xlab = "Weight (1000 lbs)",
            ylab = "P(manual transmission)")
```

## What the y axis means

This is the section worth reading before you put a `fancyfx` figure in a paper.

The two backends compute **different quantities**, and the package does not
paper over the difference:

| Model | `scale = "auto"` gives you | Centered on zero? | Other predictors |
|---|---|---|---|
| mgcv GAM | Partial effect, link scale | Yes | Excluded |
| Everything else | Predicted value, response scale | No | Held at representative values |

A **partial effect** is one term's contribution in isolation. It is centered by
construction, so its zero line is an arbitrary reference point, not a
meaningful value of the outcome. A **predicted value** is the model's actual
fitted output as one predictor varies while the others sit at their means (for
numeric predictors) or modes (for factors).

Because they are different quantities, `fancyfx` labels the y axis with whichever
one it computed, and you should not compare them side by side as though they
were on the same footing.

A GAM asked for `scale = "response"` gets *predictions*, not a back-transformed
partial effect. That is deliberate: a centered partial effect has no coherent
back-transformation on its own, and exponentiating one would produce a curve
that looks meaningful and is not.

```{r quantities}
partial <- effect_estimates(gam.fit, "Sepal.Length")
predicted <- effect_estimates(gam.fit, "Sepal.Length", scale = "response")

c(partial = attr(partial, "quantity"),
  predicted = attr(predicted, "quantity"))

# The partial effect averages to zero; the prediction does not.
round(c(partial = mean(partial$.estimate),
        predicted = mean(predicted$.estimate)), 3)
```

### The ribbon

The default `interval = "se"` draws a `± 1` standard error band, which spans
roughly 68%. This is the historical default of this package, kept so existing
figures do not silently change. For a conventional 95% interval, ask for one:

```{r intervals}
plotEffects(lm.fit, mtcars, "wt", interval = "ci", level = 0.95)
```

On the response scale, `interval = "ci"` builds the interval on the link scale
and back-transforms it, so it stays inside the range the response actually
admits — a probability band will not run past 0 or 1. That construction is
asymmetric about the estimate and so has no single standard error behind it,
which is why `interval = "se"` cannot use it and may produce a band that does
escape those bounds. If you are plotting probabilities, prefer `"ci"`.

Under a Bayesian fit, `marginaleffects` returns a credible interval, and
`interval = "ci"` will give you that rather than a confidence interval — the
argument names the ribbon, not the inferential philosophy.

## Transforms and rugs

`transform` is applied to both the curve and the rug, so the two stay on the
same x axis. Applying it to only one of them would defeat the entire purpose of
stacking them.

```{r transform}
plotEffects(lm.fit, mtcars, "hp",
            transform = "log10",
            xlab = "Horsepower (log10)")
```

Rugs come in two styles. `"histogram"` (the default) shows counts and reads well
for moderate sample sizes; `"density"` is smoother and works better when the
data are dense enough that a histogram becomes noisy.

```{r rug-types}
plotRugs(mtcars, "wt", type = "density")
```

`combinePlots()` accepts one transform for all variables or one per variable:

```{r per-var-transform}
combinePlots(lm.fit, mtcars, vars = c("wt", "hp"),
             var.transform = c("none", "log10"))
```

## Migrating from fancygam

`fancyfx` was called `fancygam` when it only handled GAMs. `plotSmooths()` still
works and produces an identical plot, but warns once per session:

```r
# Old:
plotSmooths(gam.fit, iris, "Sepal.Length")

# New — same arguments, same output:
plotEffects(gam.fit, iris, "Sepal.Length")
```

The defaults (`scale = "auto"`, `interval = "se"`) reproduce exactly what
`plotSmooths()` drew for a GAM, so migrating is a rename and nothing more.

## How to cite

```r
citation("fancyfx")
```

`fancyfx` is a plotting layer over other people's estimation work. Please also
cite whichever package actually computed your effects: **gratia** and **mgcv**
for GAM partial effects, **marginaleffects** otherwise.

## References
