---
title: "Projecting a model in space"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Projecting a model in space}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
has_terra <- requireNamespace("terra", quietly = TRUE) &&
  requireNamespace("mgcv", quietly = TRUE)
knitr::opts_chunk$set(eval = has_terra)
```

```{r setup}
library(fancyfx)
```

A projection map is a persuasive object. It fills the whole study area with
colour, it looks the same whether the model had a thousand observations there
or none, and nothing on it distinguishes the part built on evidence from the
part built on the model's willingness to keep predicting.

The two maps here are for putting that distinction back.

`terra` is a suggested package, not a required one — if you never project a
model, nothing in this vignette gets installed on your behalf.

## A worked example

A small domain with real structure: temperature falls to the north, depth
increases offshore, and — the important part — the survey only covers the
north-west.

```{r data}
library(terra)

set.seed(1)
grid <- rast(nrows = 60, ncols = 80, xmin = -71, xmax = -65,
             ymin = 41, ymax = 45, crs = "EPSG:4326")

lon <- init(grid, "x")
lat <- init(grid, "y")

sst <- 14 - 1.2 * (lat - 41) + 0.3 * (lon + 71)
names(sst) <- "sst"
depth <- 20 + 30 * (lon + 71) + 25 * (45 - lat)
names(depth) <- "depth"
covariates <- c(sst, depth)

# The survey: north-west only.
points <- data.frame(lon = runif(400, -71, -67.5), lat = runif(400, 42, 45))
survey <- cbind(points, extract(covariates, points, ID = FALSE))
survey$present <- rbinom(400, 1,
                         plogis(-4 + 0.45 * survey$sst - 0.01 * survey$depth))

sdm <- mgcv::gam(present ~ s(sst) + s(depth), data = survey,
                 family = binomial)
```

## Where does the ensemble disagree?

A single projection says what one model expects. An ensemble — competing model
formulations, emissions scenarios, bootstrap replicates — also says how much
that expectation depends on choices you could have made differently.

```{r ensemble}
members <- lapply(1:6, function(i) {
  resample <- survey[sample(nrow(survey), replace = TRUE), ]
  refit <- mgcv::gam(present ~ s(sst) + s(depth), data = resample,
                     family = binomial)
  terra::predict(covariates, refit, type = "response")
})

ensemble <- rast(members)
names(ensemble) <- paste0("bootstrap", 1:6)

plotUncertainty(ensemble, title = "Where the ensemble disagrees")
```

`statistic` chooses what "disagreement" means. `"sd"` is the default. `"cv"`
divides by the mean, which is the right choice on a count or density scale
where a spread of 5 means something very different at a mean of 10 than at
1000 — and the wrong choice for probabilities, where the mean can approach zero
and the ratio explodes.

Uncertainty is a magnitude, so it gets a sequential, perceptually uniform
viridis scale. A rainbow would invent boundaries where the data has none, which
on an uncertainty map means inventing places the ensemble agreed.

### One default worth knowing about

`na.rm` is `FALSE`, which is the opposite of most R summaries and is
deliberate. If one member is missing over part of the domain, summarising the
members that remain reports a **narrower** uncertainty exactly where the
ensemble is least complete, and nothing on the map would say so. Leaving those
cells blank makes the gap visible.

## Where is it extrapolating?

Disagreement is not the same question as novelty. Ensemble members can agree
perfectly and all be wrong together, which is precisely what happens when they
are all extrapolating into conditions none of them ever saw.

```{r mess}
plotExtrapolation(covariates, survey,
                  title = "Where the projection leaves the surveyed envelope")
```

This is a MESS surface [@elith2010]. For each cell it asks how similar the
covariates are to the values the model was fitted on, and reports the
**minimum** across covariates — one novel variable being enough to invalidate a
prediction. Below zero, the cell is outside the training range and the model is
extrapolating.

The scale diverges about zero because zero is a real boundary rather than a
midpoint of convenience: on one side the model interpolates, on the other it
guesses. The poles are red and blue rather than red and green, so the
distinction survives the common colour vision deficiencies.

Compare the two maps. The south-east is both where the ensemble disagrees most
*and* where the projection is novel — the honest reading of that region is that
the model has nothing to say about it.

To see only the part that matters:

```{r mess-novel}
plotExtrapolation(covariates, survey, novel.only = TRUE,
                  title = "Novel conditions only")
```

`training` also accepts the model itself, which saves passing the data frame
around:

```{r mess-from-model, eval = FALSE}
plotExtrapolation(covariates, sdm)
```

### What MESS does not detect

It works one covariate at a time. A cell can be perfectly ordinary on every
axis separately and still be somewhere the model has never been — warm *and*
deep, when the survey saw warm-and-shallow and cold-and-deep but never both
together. MESS reports that cell as similar.

Treat a non-negative surface as the absence of one specific problem, not as
permission to project.

## Large rasters

Real projection rasters run to millions of cells, and drawing one cell per
pixel is both slow and pointless at figure size. Above `max.cells` the raster is
aggregated first — and the plot says so in its subtitle, because silently
changing what is on the page would misrepresent the map:

```{r downsample}
plotUncertainty(ensemble, max.cells = 500)
```

## Binning into hexagons

A raster drawn at native resolution invites the reader to believe every pixel
was separately estimated, which is rarely true when the covariates were
interpolated from far coarser data. Aggregating to a cell size you can defend is
more honest than drawing detail the model does not have.

`hex_bin()` and `plotHexbin()` take either a raster or a data frame of points:

```{r hexbin}
plotHexbin(ensemble_summary(ensemble, "mean"), bins = 20,
           title = "Mean projection, binned", legend.lab = "P(present)")
```

Hexagons rather than squares for a reason worth stating: every neighbour of a
hexagon shares an edge and lies the same distance away, where a square grid has
neighbours at two different distances depending on whether they meet at an edge
or a corner. That makes hexagons better behaved for anything depending on
adjacency, and removes the visual grain a square lattice imposes.

For point data — survey effort, catch, sightings — pass the data frame and say
which column to summarise, or omit it to count:

```{r hexbin-points, fig.height = 4}
ggpubr::ggarrange(
  plotHexbin(survey, fun = "count", bins = 12, coords = c("lon", "lat"),
             title = "Survey effort", legend.lab = "Observations"),
  plotHexbin(survey, value = "present", bins = 12, coords = c("lon", "lat"),
             title = "Proportion present", legend.lab = "Proportion"),
  labels = c("A", "B")
)
```

`min.n` drops hexagons holding too few values to summarise — one observation is
not a summary of anything, and on a map it is indistinguishable from a thousand.
The returned `.n` column reports the count either way.

One caveat the function states itself, once per session: binning unprojected
longitude and latitude does not give equal-area hexagons, because a degree of
longitude shortens toward the poles. A count per hexagon is then not a density.
Project first if the areas matter.

## The numbers underneath

Both maps have a computational half that returns a raster, if you would rather
compose the figure yourself or write the result out:

```{r rasters}
spread <- ensemble_summary(ensemble, "sd")
novelty <- mess(covariates, survey)

# How much of the domain is the model extrapolating over?
mean(values(novelty) < 0, na.rm = TRUE)
```

## References
